Probing heavy ion collisions using quark and gluon jet substructure

### 2018-04-21 qg_ML.tex
YT updated the draft for JHEP submission

\abstract{
We study the phenomenon of jet quenching utilizing quark and gluon jet substructures as independent probes of heavy ion collisions. We exploit jet and subjet features to highlight differences between quark and gluon jets in vacuum and in a medium with the jet-quenching model implemented in \textsc{Jewel}. We begin with a physics-motivated, multivariate analysis of jet substructure observables including the jet mass, the radial moments, the $p_T^D$ and the pixel multiplicity. In comparison, we employ state-of-the-art image-recognition techniques by training a deep convolutional neutral network on jet images. To systematically extract jet substructure information, we introduce the telescoping deconstruction framework exploiting subjet kinematics at multiple angular scales. We draw connections to the soft-drop subjet distribution and illuminate medium-induced jet modifications using Lund diagrams. We find that the quark gluon discrimination performance worsens in heavy ion jets due to significant soft event activity affecting the soft jet substructure. Our work suggests a systematically improvable framework for studying modifications to quark and gluon jet substructures and facilitating direct comparisons between theoretical calculations, simulations and measurements in heavy ion collisions.
}

@@

Jet substructure measurements provide concrete and consistent physics information of the modification of jets in a heavy ion environment. From the measurements of the jet shape \cite{Ellis:1992qq,Chatrchyan:2013kwa,Khachatryan:2016tfj,Seymour:1997kj,Li:2011hy,Li:2012bw,Vitev:2008rz,Vitev:2009rd} and jet fragmentation function \cite{Procura:2009vm,Chatrchyan:2012gw,Aad:2014wha,Chatrchyan:2014ava,Aaboud:2017bzv,Sirunyan:2018qec}, the community has recently established that both transverse and longitudinal momentum distributions inside jets are significantly modified. Since jet substructures depend strongly on the partonic origin of jets \cite{Gallicchio:2011xq,Gallicchio:2012ez,Chien:2012ur,Chien:2015ctp,Chien:2014nsa}, a change in the fraction of quark-initiated jets and gluon-initiated jets at the LHC may significantly contribute to substructure modifications observed in experiments. It has recently been realized that an increase of the quark jet fraction in heavy ion collisions can contribute to the narrowing of jet cores as suggested in various inclusive jet substructure measurements~\cite{Chien:2015hda,Spousta:2015fca}. With jet samples consisting of different quark and gluon jet fractions, for example comparing inclusive jet and photon-tagged jet measurements, one can disentangle the effect on jet substructure modifications due to the changes in quark and gluon jet fractions caused by jet-medium interactions. This then allows one to focus on studying how quark and gluon jets are differently modified and use them as independent jet quenching probes. Note that the above jet modification picture implies that, with the high-purity quark jets in photon-tagged jet samples, the narrowing effect of jet cores should decrease. On the other hand, certain jet quenching models predict a universal narrowing of jet cores \cite{KunnawalkamElayavalli:2017hxo,Milhano:2017nzm,Casalderrey-Solana:2016jvj,Brewer:2017fqy} and emphasize contributions from medium responses \cite{Tachibana:2017syd}. Selecting purer samples of quark and gluon jets will provide the opportunity for directly studying how the medium responds to different hard probes.

@@

\begin{figure}[t]
	   \centering
	   \includegraphics[width=0.9\textwidth]{Fig12}
	   \caption{Lund diagrams for branches in the hard branch after the soft-drop procedure with $z<z_{\rm cut}=0.1$ for quark (left) and gluon (right) jets in pp (top) and central PbPb collisions in \textsc{Jewel}. See text for the definition of the axes.}
\label{fig:Lund_hard}
\end{figure}

@@

\section*{Acknowledgments}
The authors are grateful to Patrick Komiske and Eric Metodiev for collaborating on the development of telescoping deconstruction and contributions to the Appendix. The authors also thank Fr\'ed\'eric Dreyer, Benjamin Nachman, Joern Putschke, Jesse Thaler, Konrad Tywoniuk and Sevil Salur for helpful discussions. We thank Gavin Salam and the participants of the CERN TH institute ``{\sl Novel tools and observables for jet physics in heavy ion collisions}" and the 5-th Heavy Ion Jet Workshop for bringing to our attention the Lund Diagram and its usage in studying jet modifications. RKE thanks the organizers of the DS@HEP 2017 workshop at FermiLab for an excellent introduction on machine learning techniques. YTC also thanks the organizers of the BOOST2015 conference where the ideas of telescoping deconstruction and $\delta m$ were first presented. YTC was supported by the LHC Theory Initiative Postdoctoral Fellowship under the National Science Foundation grant PHY-1419008. RKE also acknowledges support in part from the National Science Foundation under Grant No.1067907 \& 1352081 and in part by the U.S. Department of Energy Office of Science, Office of Nuclear Physics under Award Number DE-FG02-92ER-40713.

>>>>

\abstract{
We study the phenomenon of jet quenching utilizing quark and gluon jet substructures as independent probes of heavy ion collisions. We exploit jet and subjet features to highlight differences between quark and gluon jets in vacuum and in a medium with the jet-quenching model implemented in \textsc{Jewel}. We begin with a physics-motivated, multivariate analysis of jet substructure observables including the jet mass, the radial moments, the $p_T^D$ and the pixel multiplicity. In comparison, we employ state-of-the-art image-recognition techniques by training a deep convolutional neural network on jet images. To systematically extract jet substructure information, we introduce the telescoping deconstruction framework exploiting subjet kinematics at multiple angular scales. We draw connections to the soft-drop subjet distribution and illuminate medium-induced jet modifications using Lund diagrams. We find that the quark gluon discrimination performance worsens in heavy ion jets due to significant soft event activity affecting the soft jet substructure. Our work suggests a systematically improvable framework for studying modifications to quark and gluon jet substructures and facilitating direct comparisons between theoretical calculations, simulations and measurements in heavy ion collisions.
}

@@

Jet substructure measurements provide concrete and consistent physics information of the modification of jets in a heavy ion environment. From the measurements of the jet shape \cite{Ellis:1992qq,Chatrchyan:2013kwa,Khachatryan:2016tfj,Seymour:1997kj,Li:2011hy,Li:2012bw,Vitev:2008rz,Vitev:2009rd} and jet fragmentation function \cite{Procura:2009vm,Chatrchyan:2012gw,Aad:2014wha,Chatrchyan:2014ava,Aaboud:2017bzv,Sirunyan:2018qec}, the community has recently established that both transverse and longitudinal momentum distributions inside jets are significantly modified. Since jet substructures depend strongly on the partonic origin of jets \cite{Gallicchio:2011xq,Gallicchio:2012ez,Chien:2012ur,Chien:2015ctp,Chien:2014nsa}, a change in the fraction of quark-initiated jets and gluon-initiated jets from proton-proton to heavy ion collisions may significantly contribute to substructure modifications observed in experiments. It has recently been realized that an increase of the quark jet fraction in heavy ion collisions can contribute to the narrowing of jet cores as suggested in various inclusive jet substructure measurements~\cite{Chien:2015hda,Spousta:2015fca}. With jet samples consisting of different quark and gluon jet fractions, for example comparing inclusive jet and photon-tagged jet measurements, one can disentangle the effect on jet substructure modifications due to the changes in quark and gluon jet fractions caused by jet-medium interactions. This then allows one to focus on studying how quark and gluon jets are differently modified and use them as independent jet quenching probes. Note that the above jet modification picture implies that, with the high-purity quark jets in photon-tagged jet samples, the narrowing effect of jet cores should decrease. On the other hand, certain jet quenching models predict a universal narrowing of jet cores \cite{KunnawalkamElayavalli:2017hxo,Milhano:2017nzm,Casalderrey-Solana:2016jvj,Brewer:2017fqy} and emphasize contributions from medium responses \cite{Tachibana:2017syd}. Selecting purer samples of quark and gluon jets will provide the opportunity for directly studying how the medium responds to different hard probes.

@@

\begin{figure}[t]
	   \centering
	   \includegraphics[width=0.9\textwidth]{Fig12}
	   \caption{Lund diagrams for branches in the hard branch ($z>z_{\rm cut}=0.1$) after the soft-drop procedure for quark (left) and gluon (right) jets in pp (top) and central PbPb collisions in \textsc{Jewel}. Note the significant fraction of branches with $z<z_{\rm cut}$ and the increase of branches with $z\gtrsim z_{\rm cut}$ at wide angles in PbPb collisions. See text for the definition of the axes.}
\label{fig:Lund_hard}
\end{figure}

@@

\section*{Acknowledgments}
The authors are grateful to Patrick Komiske and Eric Metodiev for collaborating on the development of telescoping deconstruction and contributions to the Appendix. The authors also thank Fr\'ed\'eric Dreyer, Benjamin Nachman, Joern Putschke, Jesse Thaler, Konrad Tywoniuk and Sevil Salur for helpful discussions and comments. We thank Gavin Salam and the participants of the CERN TH institute ``{\sl Novel tools and observables for jet physics in heavy ion collisions}" and the 5-th Heavy Ion Jet Workshop for bringing to our attention the Lund Diagram and its usage in studying jet modifications. RKE thanks the organizers of the DS@HEP 2017 workshop at FermiLab for an excellent introduction on machine learning techniques. YTC also thanks the organizers of the BOOST2015 conference where the ideas of telescoping deconstruction and $\delta m$ were first presented. YTC was supported by the LHC Theory Initiative Postdoctoral Fellowship under the National Science Foundation grant PHY-1419008. RKE also acknowledges support in part from the National Science Foundation under Grant No.1067907 \& 1352081 and in part by the U.S. Department of Energy Office of Science, Office of Nuclear Physics under Award Number DE-FG02-92ER-40713.
